---
title: "Data transformation with dplyr"
author: "John Blischak"
date: "2019-09-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/transform.html)
* [Source](https://github.com/hadley/r4ds/blob/master/transform.Rmd)

## Setup

```{r}
library(nycflights13)
library(tidyverse)
```

## Filter rows with `filter()`

p. 49

> Find all flights that

> Had an arrival delay of two or more hours
    
```{r}
filter(flights, arr_delay > 2 * 60)
```
    
> Flew to Houston (`IAH` or `HOU`)
    
```{r}
filter(flights, dest == "IAH" | dest == "HOU")
```

> Were operated by United, American, or Delta
    
```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

> Departed in summer (July, August, and September)
        
```{r}
filter(flights, month %in% 7:9)
```

> Arrived more than two hours late, but didn't leave late
        
```{r}
filter(flights, arr_delay > 2 * 60, dep_delay <=0)
```

> Were delayed by at least an hour, but made up over 30 minutes in flight
        
```{r}
filter(flights, dep_delay >= 60, arr_delay < 30)
```

> Departed between midnight and 6am (inclusive)
    
```{r}
filter(flights, hour <= 6)
```

> Another useful dplyr filtering helper is `between()`. What does it do?
    Can you use it to simplify the code needed to answer the previous 
    challenges?

`between()` selects numeric values between a minimum and maximum value (inclusive).

While I think `between()` is useful, I don't find these examples very compelling.

```{r}
filter(flights, between(month, 7, 9))
filter(flights, between(hour, 0, 6))
```

> How many flights have a missing `dep_time`? What other variables are 
    missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

I assume these are cancelled flights since all columns relating to departure and
arrival times are missing.

> Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing?
    Why is `FALSE & NA` not missing? Can you figure out the general
    rule?  (`NA * 0` is a tricky counterexample!)

```{r}
NA ^ 0 # same as NA ^ FALSE
NA | TRUE
FALSE & NA
NA * 0
```

I think the `NA | TRUE` and `FALSE & NA` make sense. They are short-circuitng
the logic. For an "or" statement, a `TRUE` on one of the sides is sufficient to
render the entire statement true, regardless if the other data is missing. Vice
verse for the "and": a `FALSE` on either side is sufficient to make the entire
statement false.

```{r}
# have to be TRUE
NA | TRUE
TRUE | NA
# ambiguous
NA | FALSE
FALSE | NA
```

```{r}
# have to be TRUE
NA & FALSE
FALSE & NA
# ambiguous
NA & TRUE
TRUE & NA
```

The exponentiation logic also makes less sense to me. Any number raised to the
0th power is equal to 1. Furthermore, 1 raised to any power is also 1.
From the docs for `?Arithmetic`:

> 1 ^ y and y ^ 0 are 1, *always*.

```{r}
NA ^ 0
1 ^ NA
```

Thus the general rule seems to be to allow `NA` if the missing value would not
affect the outcome of the logic or arithmetic operation.

Thus I'm not sure why `NA * 0`  returns `NA` when it should be `0` no matter the
value. My only guess is that it may be related to the potential risk of division
by zero.

```{r}
NA * 0
0 / NA
NA / NA * 0
```

## Arrange rows with `arrange()`

p. 51

> How could you use `arrange()` to sort all missing values to the start?
>   (Hint: use `is.na()`).

```{r}
arrange(flights, desc(is.na(dep_time)))
```

> Sort `flights` to find the most delayed flights. Find the flights that
    left earliest.

```{r}
arrange(flights, desc(dep_delay))
arrange(flights, hour, minute) # left earliest in the AM
arrange(flights, dep_delay) # left earliest in relation to scheduled dep time
```

> Sort `flights` to find the fastest flights.

```{r}
arrange(flights, air_time)
```

> Which flights travelled the longest? Which travelled the shortest?

```{r}
arrange(flights, desc(distance))
arrange(flights, distance)
```

## Select columns with `select()`

p. 54

> Brainstorm as many ways as possible to select `dep_time`, `dep_delay`,
    `arr_time`, and `arr_delay` from `flights`.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, dep_time, dep_delay:arr_time, arr_delay)
select(flights, starts_with("dep"), starts_with("arr"))
select(flights, dep_time, arr_time, ends_with("delay"))
```

> What happens if you include the name of a variable multiple times in
    a `select()` call?

The column is added in the first place it is mentioned:

```{r}
select(flights, dep_time, dep_time)
select(flights, dep_time, arr_time, dep_time)
```

> What does the `one_of()` function do? Why might it be helpful in conjunction
    with this vector?

From docs:

> `one_of()`: Matches variable names in a character vector.

Thus it serves a similar purpose as `%in%`. But `%in%` returns a logical vector (which could be used in `filter()`), whereas, `select()` accepts the integer position of the columns:

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
# This would be how to replace one_of()
select(flights, which(colnames(flights) %in% vars))
```

> Does the result of running the following code surprise you?  How do the
    select helpers deal with case by default? How can you change that default?

```{r}
select(flights, contains("TIME"))
```

The tidyselect helpers ignore case by default (`ignore.case = TRUE`):

```{r}
formals(contains)$ignore.case
select(flights, contains("TIME", ignore.case = FALSE))
```

## Add new variables with `mutate()`

p. 58

> Currently `dep_time` and `sched_dep_time` are convenient to look at, but
    hard to compute with because they're not really continuous numbers. 
    Convert them to a more convenient representation of number of minutes
    since midnight.

```{r}
select(flights, dep_time, sched_dep_time)
transmute(flights, dep_time, sched_dep_time,
          dep_time_min = dep_time %/% 100 * 60 + dep_time %% 100,
          sched_dep_time_min = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100)
```

> Compare `air_time` with `arr_time - dep_time`. What do you expect to see?
    What do you see? What do you need to do to fix it?

I expect them to be the same.

```{r}
transmute(flights, air_time, arr_time - dep_time)
```

`arr_time - dep_time` is greater than `air_time`

```{r}
transmute(flights, air_time, total = arr_time - dep_time, total - air_time)
```

`air_time` is in minutes, but the delays are in HHMM format. I first need to
convert the delays to minutes from midnight, as above. I am pretty sure the
departure and arrival times are reported using the same timezone.

```{r}
minutes <- transmute(flights,
                     dep_time,
                     dep_time_min = dep_time %/% 100 * 60 + dep_time %% 100,
                     arr_time,
                     arr_time_min = arr_time %/% 100 * 60 + arr_time %% 100,
                     air_time,
                     total = arr_time_min - dep_time_min,
                     concord = air_time == total)
minutes
```

I checked to see if `air_time` was computed from the scheduled departure and
arrival times, but these didn't match either.

```{r}
minutes <- transmute(flights,
                     sched_dep_time,
                     sched_dep_time_min = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100,
                     sched_arr_time,
                     sched_arr_time_min = sched_arr_time %/% 100 * 60 + sched_arr_time %% 100,
                     air_time,
                     total = sched_arr_time_min - sched_dep_time_min,
                     concord = air_time == total)
minutes
```

Doing some research online
([source](https://jrnold.github.io/r4ds-exercise-solutions/transform.html#add-new-variables-with-mutate)),
it appears that these won't match up since `air_time` does not include taxi-ing.

> Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you
    expect those three numbers to be related?

```{r}
transmute(flights, dep_time, sched_dep_time, dep_delay)
```

I expect `dep_delay` to be the number of minutes between `dep_time` and
`sched_dep_time`.

```{r}
transmute(flights, dep_time, sched_dep_time, dep_delay,
          dep_time_min = dep_time %/% 100 * 60 + dep_time %% 100,
          sched_dep_time_min = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100,
          expected_delay = dep_time_min - sched_dep_time_min,
          concond = expected_delay == dep_delay)
```

> Find the 10 most delayed flights using a ranking function. How do you want 
    to handle ties? Carefully read the documentation for `min_rank()`.

Use `desc()` with `min_rank()` so that the most delayed flight is ranked first.

```{r}
delayed <- transmute(flights, carrier, flight, dep_delay, rank = min_rank(desc(dep_delay)))
arrange(delayed, rank)
```

The `NA`'s are left as is:

```{r}
sum(is.na(delayed$dep_delay))
sum(is.na(delayed$rank))
```

> What does `1:3 + 1:10` return? Why?

```{r}
1:3 + 1:10
```

The `1:3` is recycled. Thus it gets used 3 times, and then just the final `1`.
Thus it is equivalent to the following:

```{r}
c(1:3 + 1:3, 1:3 + 4:6, 1:3 + 7:9, 1 + 10)
```

> What trigonometric functions does R provide?

```{r}
?Trig
```

```
cos(x)
sin(x)
tan(x)

acos(x)
asin(x)
atan(x)
atan2(y, x)

cospi(x)
sinpi(x)
tanpi(x)
```

## Group summaries with `summarize()`

p. 72

> Brainstorm at least 5 different ways to assess the typical delay 
    characteristics of a group of flights. Consider the following scenarios:
    
    * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of 
      the time.
      
    * A flight is always 10 minutes late.

    * A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of 
      the time.
      
    * 99% of the time a flight is on time. 1% of the time it's 2 hours late.
    
    Which is more important: arrival delay or departure delay?

> Come up with another approach that will give you the same output as 
    `not_cancelled %>% count(dest)` and 
    `not_cancelled %>% count(tailnum, wt = distance)` (without using 
    `count()`).

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled %>% count(dest)
not_cancelled %>%
  group_by(dest) %>%
  summarize(n = n())
not_cancelled %>% count(tailnum, wt = distance)
not_cancelled %>%
  group_by(tailnum) %>%
  summarize(n = sum(distance))
```

> Our definition of cancelled flights (`is.na(dep_delay) | is.na(arr_delay)`
    ) is slightly suboptimal. Why? Which is the most important column?

I think `arr_delay` is more important. If a flight takes off on time, but
lands at a different airport, that's a big issue.

There are 1,175 flights that have a `dep_delay` but not a `arr_delay`.

```{r}
sum(is.na(flights$dep_delay) | is.na(flights$arr_delay))
sum(is.na(flights$dep_delay))
sum(is.na(flights$arr_delay))
sum(xor(is.na(flights$dep_delay), is.na(flights$arr_delay)))
sum(is.na(flights$dep_delay) & !is.na(flights$arr_delay))
sum(!is.na(flights$dep_delay) & is.na(flights$arr_delay))
```

Thus I guess a more strict definintion of "cancelled" would be `is.na(dep_delay)`
and the few cases where the plane departs but doesn't arrive could be classified
as "no arrivals" (presumably this includes redirections and crashes).

> Look at the number of cancelled flights per day. Is there a pattern?
    Is the proportion of cancelled flights related to the average delay?

```{r}
flights %>%
  mutate(cancelled = is.na(dep_delay)) %>%
  group_by(year, month, day) %>%
  summarize(n = n(),
            prop_canc = sum(cancelled) / n,
            avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = avg_delay, y = prop_canc)) +
  geom_point(alpha = 0.25) +
  geom_smooth()
```

There is a slight pattern, but it's mostly driven by the days where the average
delay is greater than 10 minutes.

> Which carrier has the worst delays? Challenge: can you disentangle the
    effects of bad airports vs. bad carriers? Why/why not? (Hint: think about
    `flights %>% group_by(carrier, dest) %>% summarise(n())`)

```{r}
flights %>%
  group_by(carrier) %>%
  summarize(mean_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(mean_delay))
```

```{r}
flights %>% group_by(carrier, dest) %>% summarise(n())
```

> What does the `sort` argument to `count()` do. When might you use it?

Sorts by value of `n`. Can be used instead of piping to `arrange()`.

## Group mutates (and filters)

> A grouped filter is a grouped mutate followed by an ungrouped filter.

Group filters are miserable to try and reason about. I can't imagine ever
wanting to use them. I find it much more natural to use `summarize()`.

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
jan_1 <- flights_sml %>%
  filter(year == 2013, month == 1, day == 1)
dim(jan_1)
desc(jan_1$arr_delay)
rank(desc(jan_1$arr_delay))
rank(desc(jan_1$arr_delay)) < 10
# OK. Starting to make sense. For each day, it keeps the 9 worst flights
```

```{r}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
popular_dests
length(unique(flights$dest))
length(unique(popular_dests$dest))
```

p. 75

> Refer back to the lists of useful mutate and filtering functions. 
    Describe how each operation changes when you combine it with grouping.

> Which plane (`tailnum`) has the worst on-time record?

```{r}
# interpretation: worst arrival delay for flights that arrived at their destination
flights %>%
  filter(!is.na(tailnum), !is.na(arr_delay)) %>%
  group_by(tailnum) %>%
  summarize(mean_delay = mean(arr_delay)) %>%
  filter(rank(desc(mean_delay)) < 10)
```

> What time of day should you fly if you want to avoid delays as much
    as possible?

You would want to leave before 9 AM. Solution below using `summarize()` +
`arrange()`. I assume you can also do this with only `filter()`, but I don't
see the advantage.

```{r}
not_cancelled %>%
  group_by(hour) %>%
  summarize(mean_delay = mean(arr_delay)) %>%
  arrange(mean_delay)
```

> For each destination, compute the total minutes of delay. For each 
    flight, compute the proportion of the total delay for its destination.

Total minutes of delay per destination

```{r}
not_cancelled %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  summarize(total_delay = sum(arr_delay))
```

Per flight proportion of delay per destination

```{r}
props <- not_cancelled %>%
  filter(arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(flight, dest, arr_delay, prop_delay)
props
sum(props$prop_delay[props$dest == "IAH"])
```

> Delays are typically temporally correlated: even once the problem that
    caused the initial delay has been resolved, later flights are delayed 
    to allow earlier flights to leave. Using `lag()`, explore how the delay
    of a flight is related to the delay of the immediately preceding flight.

> Look at each destination. Can you find flights that are suspiciously
    fast? (i.e. flights that represent a potential data entry error). Compute
    the air time a flight relative to the shortest flight to that destination.
    Which flights were most delayed in the air?
    
> Find all destinations that are flown by at least two carriers. Use that
    information to rank the carriers.

> For each plane, count the number of flights before the first delay 
    of greater than 1 hour.
