---
title: "Data transformation with dplyr"
author: "John Blischak"
date: "2019-09-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/transform.html)
* [Source](https://github.com/hadley/r4ds/blob/master/transform.Rmd)

## Setup

```{r}
library(nycflights13)
library(tidyverse)
```

## Filter rows with `filter()`

p. 49

> Find all flights that

> Had an arrival delay of two or more hours
    
```{r}
filter(flights, arr_delay > 2 * 60)
```
    
> Flew to Houston (`IAH` or `HOU`)
    
```{r}
filter(flights, dest == "IAH" | dest == "HOU")
```

> Were operated by United, American, or Delta
    
```{r}
filter(flights, carrier %in% c("UA", "AA", "DL"))
```

> Departed in summer (July, August, and September)
        
```{r}
filter(flights, month %in% 7:9)
```

> Arrived more than two hours late, but didn't leave late
        
```{r}
filter(flights, arr_delay > 2 * 60, dep_delay <=0)
```

> Were delayed by at least an hour, but made up over 30 minutes in flight
        
```{r}
filter(flights, dep_delay >= 60, arr_delay < 30)
```

> Departed between midnight and 6am (inclusive)
    
```{r}
filter(flights, hour <= 6)
```

> Another useful dplyr filtering helper is `between()`. What does it do?
    Can you use it to simplify the code needed to answer the previous 
    challenges?

`between()` selects numeric values between a minimum and maximum value (inclusive).

While I think `between()` is useful, I don't find these examples very compelling.

```{r}
filter(flights, between(month, 7, 9))
filter(flights, between(hour, 0, 6))
```

> How many flights have a missing `dep_time`? What other variables are 
    missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

I assume these are cancelled flights since all columns relating to departure and
arrival times are missing.

> Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing?
    Why is `FALSE & NA` not missing? Can you figure out the general
    rule?  (`NA * 0` is a tricky counterexample!)

```{r}
NA ^ 0 # same as NA ^ FALSE
NA | TRUE
FALSE & NA
NA * 0
```

I think the `NA | TRUE` and `FALSE & NA` make sense. They are short-circuitng
the logic. For an "or" statement, a `TRUE` on one of the sides is sufficient to
render the entire statement true, regardless if the other data is missing. Vice
verse for the "and": a `FALSE` on either side is sufficient to make the entire
statement false.

```{r}
# have to be TRUE
NA | TRUE
TRUE | NA
# ambiguous
NA | FALSE
FALSE | NA
```

```{r}
# have to be TRUE
NA & FALSE
FALSE & NA
# ambiguous
NA & TRUE
TRUE & NA
```

The "exclusive or" logic makes less sense to me. From the docs for `?Arithmetic`:

> 1 ^ y and y ^ 0 are 1, *always*.

And that doesn't make sense to me.

```{r}
NA ^ FALSE
NA ^ 1
1 ^ NA
1 ^ 1
0 ^ 0
```

While R is willing to make assumptions about logic statements, it doesn't do
this for arithmetic, e.g. `NA * 0` is likely `0` no matter the value. I assume
this is riskier due to potential division by zero.

```{r}
0 / NA
NA / NA * 0
```

## Arrange rows with `arrange()`

p. 51

> How could you use `arrange()` to sort all missing values to the start?
>   (Hint: use `is.na()`).

```{r}
arrange(flights, desc(is.na(dep_time)))
```

> Sort `flights` to find the most delayed flights. Find the flights that
    left earliest.

```{r}
arrange(flights, desc(dep_delay))
arrange(flights, hour, minute) # left earliest in the AM
arrange(flights, dep_delay) # left earliest in relation to scheduled dep time
```

> Sort `flights` to find the fastest flights.

```{r}
arrange(flights, air_time)
```

> Which flights travelled the longest? Which travelled the shortest?

```{r}
arrange(flights, desc(distance))
arrange(flights, distance)
```

## Select columns with `select()`

p. 54

> Brainstorm as many ways as possible to select `dep_time`, `dep_delay`,
    `arr_time`, and `arr_delay` from `flights`.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, dep_time, dep_delay:arr_time, arr_delay)
select(flights, starts_with("dep"), starts_with("arr"))
select(flights, dep_time, arr_time, ends_with("delay"))
```

> What happens if you include the name of a variable multiple times in
    a `select()` call?

The column is added in the first place it is mentioned:

```{r}
select(flights, dep_time, dep_time)
select(flights, dep_time, arr_time, dep_time)
```

> What does the `one_of()` function do? Why might it be helpful in conjunction
    with this vector?

From docs:

> `one_of()`: Matches variable names in a character vector.

Thus it serves a similar purpose as `%in%`. But `%in%` returns a logical vector (which could be used in `filter()`), whereas, `select()` accepts the integer position of the columns:

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
# This would be how to replace one_of()
select(flights, which(colnames(flights) %in% vars))
```

> Does the result of running the following code surprise you?  How do the
    select helpers deal with case by default? How can you change that default?

```{r}
select(flights, contains("TIME"))
```

The tidyselect helpers ignore case by default (`ignore.case = TRUE`):

```{r}
formals(contains)$ignore.case
select(flights, contains("TIME", ignore.case = FALSE))
```
