---
title: "Exploratory data analysis"
author: "John Blischak"
date: "2019-09-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/exploratory-data-analysis.html)
* [Source](https://github.com/hadley/r4ds/blob/master/EDA.Rmd)

## Setup

```{r setup, message=FALSE}
library(nycflights13)
library(tidyverse)
data(diamonds)
library(ggstance)
library(lvplot)
library(ggbeeswarm)
```

## Variation

p. 90

> Explore the distribution of each of the `x`, `y`, and `z` variables 
    in `diamonds`. What do you learn? Think about a diamond and how you
    might decide which dimension is the length, width, and depth.

```{r}
ggplot(data = diamonds, mapping = aes(x = x)) +
  geom_histogram()
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram()
ggplot(data = diamonds, mapping = aes(x = z)) +
  geom_histogram()
```

> Explore the distribution of `price`. Do you discover anything unusual
    or surprising? (Hint: Carefully think about the `binwidth` and make sure
    you try a wide range of values.)

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 10)
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 100)
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram(binwidth = 1000)
```

> How many diamonds are 0.99 carat? How many are 1 carat? What
    do you think is the cause of the difference?

```{r}
diamonds %>%
  filter(between(carat, 0.99, 1)) %>%
  count(carat)
```

> Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when
    zooming in on a histogram. What happens if you leave `binwidth` unset?
    What happens if you try and zoom so only half a bar shows?

Using `coord_cartesian()` zooms into the existing plot. Thus the bin width
does not change. Using `xlim()` changes the binwidth because it changes
the number of observations. Using `ylim()` removes data because the bars that
go above 1000 are removed from the final plot.

```{r}
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram()
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram() +
  xlim(5, 10)
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram() +
  ylim(0, 1000)
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram() +
  coord_cartesian(xlim = c(5, 10))
ggplot(data = diamonds, mapping = aes(x = y)) +
  geom_histogram() +
  coord_cartesian(ylim = c(0, 1000))
```

## Missing values

p. 93

> What happens to missing values in a histogram?  What happens to missing
    values in a bar chart? Why is there a difference?

`geom_histogram()` removes missing values numeric variables:

```{r}
d <- data.frame(x = c(1, 1, 2, 2, NA, 3))
ggplot(data = d, mapping = aes(x = x)) +
  geom_histogram()
```

`geom_bar()` keeps missing values for non-numeric variables:

```{r}
d <- data.frame(x = c("a", "a", "b", "b", NA, "c"))
ggplot(data = d, mapping = aes(x = x)) +
  geom_bar()
```

An edge case is integers. They are treated as numeric and have missing values removed.
To treat them as discrete, they need to be converted to a factor:

```{r}
d <- data.frame(x = c(1, 1, 2, 2, NA, 3))
ggplot(data = d, mapping = aes(x = x)) +
  geom_bar()
d$x <- factor(d$x)
ggplot(data = d, mapping = aes(x = x)) +
  geom_bar()
```

> What does `na.rm = TRUE` do in `mean()` and `sum()`?

Missing values are removed prior to performing the calculation.

## Covariation - A categorical and continuous variable

p. 99

> Use what you've learned to improve the visualisation of the departure times
    of cancelled vs. non-cancelled flights.

Plot the density values:

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(y = ..density.., colour = cancelled), binwidth = 1/4)
```

I prefer the version with `geom_density()`:

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_density(mapping = aes(colour = cancelled), bw = 1/4)
```

> What variable in the diamonds dataset is most important for predicting
    the price of a diamond? How is that variable correlated with cut?
    Why does the combination of those two relationships lead to lower quality
    diamonds being more expensive?

carat is the most important for predicting price.

```{r}
summary(lm(price ~ ., data = diamonds))
```

Lower quality cuts have higher carat values.

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = carat)) +
  geom_boxplot()
```

And since a higher carat gets a higher price, lower cuts tend to be more
expensive on average.

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ cut)
```

> Install the ggstance package, and create a horizontal boxplot.
    How does this compare to using `coord_flip()`?

Have to switch `x` and `y` aesthetics:

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = carat)) +
  geom_boxplot() +
  coord_flip()
ggplot(data = diamonds, mapping = aes(x = carat, y = cut)) +
  ggstance::geom_boxploth()
```

> One problem with boxplots is that they were developed in an era of 
    much smaller datasets and tend to display a prohibitively large
    number of "outlying values". One approach to remedy this problem is
    the letter value plot. Install the lvplot package, and try using
    `geom_lv()` to display the distribution of price vs cut. What
    do you learn? How do you interpret the plots?

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  lvplot::geom_lv()
```

> Compare and contrast `geom_violin()` with a facetted `geom_histogram()`,
    or a coloured `geom_freqpoly()`. What are the pros and cons of each 
    method?

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_violin()
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_histogram() +
  facet_wrap(~ cut)
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_freqpoly(mapping = aes(color = cut))
```

> If you have a small dataset, it's sometimes useful to use `geom_jitter()`
    to see the relationship between a continuous and categorical variable.
    The ggbeeswarm package provides a number of methods similar to 
    `geom_jitter()`. List them and briefly describe what each one does.

```{r}
ggplot(data = diamonds[sample(seq_len(nrow(diamonds)), 1000), ],
       mapping = aes(x = cut, y = price)) +
  ggbeeswarm::geom_beeswarm()
```

## Covariation - Two categorical variables

p. 101

> How could you rescale the count dataset above to more clearly show
    the distribution of cut within colour, or colour within cut?

Use a grouped mutate to calculate proportion of each cut within each color.

```{r}
diamonds %>% 
  count(color, cut) %>%
  group_by(color) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = prop))
```

> Use `geom_tile()` together with dplyr to explore how average flight
    delays vary by destination and month of year.  What makes the 
    plot difficult to read? How could you improve it?

Only including non-cancelled flights with delayed arrivals.

```{r}
flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay), arr_delay > 0) %>%
  select(dest, month, arr_delay) %>%
  group_by(dest, month) %>%
  summarize(delay = mean(arr_delay)) %>%
  ggplot(mapping = aes(x = as.factor(month), y = dest)) +
  geom_tile(mapping = aes(fill = delay))
```

> Why is it slightly better to use `aes(x = color, y = cut)` rather
    than `aes(x = cut, y = color)` in the example above?

Putting longer variable values on y-axis ensure that the axis labels don't
overlap.

```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = n))
diamonds %>% 
  count(color, cut) %>%  
  ggplot(mapping = aes(x = cut, y = color)) +
    geom_tile(mapping = aes(fill = n))
```

## Covariation - Two continuous variables

```{r}
smaller <- diamonds %>% 
  filter(carat < 3)
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_width(carat, 0.1)), varwidth = TRUE)
ggplot(data = smaller, mapping = aes(x = carat, y = price)) + 
  geom_boxplot(mapping = aes(group = cut_number(carat, 20)))
```

I think the above plot is confusing. The use of `cut_number()` puts a fairly
equal number of observations in each bin. However, because `carat` is mapped to
`x`, the width of the bin corresponds to the width on the x-axis, not the number
of observations (as is done with `varwidth = TRUE`).

```{r}
table(cut_number(diamonds$carat, 20))
```

p. 105

> Instead of summarising the conditional distribution with a boxplot, you
    could use a frequency polygon. What do you need to consider when using
    `cut_width()` vs `cut_number()`? How does that impact a visualisation of
    the 2d distribution of `carat` and `price`?

With `cut_width()`, the number of observations in each bin may vary substantially;
whereas, with `cut_number()`, each bin will have a similar number of observations.

```{r}
# cut_interval makes n groups with equal range
table(cut_interval(diamonds$price, n = 10))

# cut_number makes n groups with (approximately) equal numbers of observations
table(cut_number(diamonds$price, n = 10))

# cut_width makes groups of width width.
table(cut_width(diamonds$price, width = 1000))
```


```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) +
  geom_freqpoly(mapping = aes(color = cut_width(price, width = 1000)))
ggplot(data = diamonds, mapping = aes(x = carat)) +
  geom_freqpoly(mapping = aes(color = cut_number(price, n = 10)))
```

> Visualise the distribution of carat, partitioned by price.

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = carat)) +
  geom_boxplot(mapping = aes(group = cut_width(price, 1000)))
```

> How does the price distribution of very large diamonds compare to small 
    diamonds? Is it as you expect, or does it surprise you?

```{r}
diamonds %>%
  mutate(volume = x * y * z, # naive, but not clear what they want
         size = ifelse(volume > quantile(volume, 0.75), "large",  "small")) %>% 
  ggplot(mapping = aes(x = price)) +
  geom_density(mapping = aes(color = size))
```
    
> Combine two of the techniques you've learned to visualise the 
    combined distribution of cut, carat, and price.

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(mapping = aes(color = cut), alpha = 0.5)
```

1. Two dimensional plots reveal outliers that are not visible in one 
   dimensional plots. For example, some points in the plot below have an 
   unusual combination of `x` and `y` values, which makes the points outliers 
   even though their `x` and `y` values appear normal when examined separately.

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
```

>   Why is a scatterplot a better display than a binned plot for this case?

Because you want to highlight outliers. A binned plot would focus on the most
densely overplotted areas.

