---
title: "Tibbles with tibble"
author: "John Blischak"
date: "2019-09-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/tibbles.html)
* [Source](https://github.com/hadley/r4ds/blob/master/tibble.Rmd)

## Setup

```{r setup, message=FALSE}
library(tidyverse)
```

## Exercises

p. 123

> How can you tell if an object is a tibble? (Hint: try printing `mtcars`,
    which is a regular data frame).

```{r}
class(mtcars)
class(as_tibble(mtcars))
```

> Compare and contrast the following operations on a `data.frame` and 
    equivalent tibble. What is different? Why might the default data frame
    behaviours cause you frustration?

```{r}
# data frame
df <- data.frame(abc = 1, xyz = "a")
df$x # Gives you a result (column xyz) even though column x does not exist
df[, "xyz"] # returns a vector, string coerced to factor
df[, c("abc", "xyz")] # returns a data frame

# tibble
dft <- tibble(abc = 1, xyz = "a")
dft$x # warns that column x does not exist
dft[, "xyz"] # returns tibble, string remains string
dft[, c("abc", "xyz")] # returns tibble
```

> If you have the name of a variable stored in an object, e.g. `var <- "mpg"`,
    how can you extract the reference variable from a tibble?

```{r}
var <- "mpg"
head(mtcars[, var])
head(as_tibble(mtcars)[, var])
```

> Practice referring to non-syntactic names in the following data frame by:
> Extracting the variable called `1`.
> Plotting a scatterplot of `1` vs `2`.
> Creating a new column called `3` which is `2` divided by `1`.
        
> Renaming the columns to `one`, `two` and `three`. 

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
annoying$`1`
annoying[["1"]]
ggplot(annoying, aes(`1`, `2`)) + geom_point()
annoying <- annoying %>%
  mutate(`3` = `1` + `2`)
annoying %>% rename(one = `1`, two = `2`, three = `3`)
```
> What does `tibble::enframe()` do? When might you use it?

Converts a named vector/list to a 2-column data frame.

> What option controls how many additional column names are printed
    at the footer of a tibble?

> tibble.max_extra_cols: Number of extra columns printed in reduced form. Default: 100.
