---
title: "Data import with readr"
author: "John Blischak"
date: "2019-10-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/data-import.html)
* [Source](https://github.com/hadley/r4ds/blob/master/import.Rmd)

## Setup

```{r setup, message=FALSE}
library(tidyverse)
```

## Getting started

p. 128

1.  What function would you use to read a file where fields were separated with  
    "|"?

`read_delim()` with `delim = "|"`.

1.  Apart from `file`, `skip`, and `comment`, what other arguments do
    `read_csv()` and `read_tsv()` have in common?

```{r}
intersect(names(formals(read_csv)), names(formals(read_tsv)))
```

In fact they share all the same arguments:

```{r}
identical(names(formals(read_csv)), names(formals(read_tsv)))
```

`read_csv()` and `read_tsv()` are both wrappers to the internal function
`read_delimited()`:

```{r}
names(formals(readr:::read_delimited))
```

1.  What are the most important arguments to `read_fwf()`?

```{r}
names(formals(read_fwf))
```

` col_positions`

> col_positions	
Column positions, as created by fwf_empty(), fwf_widths() or fwf_positions(). To read in only selected fields, use fwf_positions(). If the width of the last column is variable (a ragged fwf file), supply the last end position as NA.

1.  Sometimes strings in a CSV file contain commas. To prevent them from
    causing problems they need to be surrounded by a quoting character, like
    `"` or `'`. By convention, `read_csv()` assumes that the quoting
    character will be `"`, and if you want to change it you'll need to
    use `read_delim()` instead. What arguments do you need to specify
    to read the following text into a data frame?
    
```{r, eval = FALSE}
"x,y\n1,'a,b'"
```

Set `quote` to `'`:

```{r}
read_delim("x,y\n1,'a,b'", delim = ",", quote = "'")
```

As of readr
[1.1.0](https://github.com/tidyverse/readr/blob/master/NEWS.md#readr-110)
(released in March 2017), you can just use `read_csv()`:

```{r}
read_csv("x,y\n1,'a,b'", quote = "'")
```

1.  Identify what is wrong with each of the following inline CSV files. 
    What happens when you run the code?
    
```{r, eval = TRUE}
# 2 column names but 3 columns of data
read_csv("a,b\n1,2,3\n4,5,6")
# Each row has a different number of columns
read_csv("a,b,c\n1,2\n1,2,3,4")
# There is an opening quote in the second row but no closing quote
read_csv("a,b\n\"1")
# Both columns are characters b/c they contain a mix of numbers and characters
read_csv("a,b\n1,2\na,b")
# The delimiter is a `;`, so everything is in one column
read_csv("a;b\n1;3")
```

## Parsing a vector

p. 136

1.  What are the most important arguments to `locale()`? 

Seems like a pretty context-dependent question. In this chapter, they use
`decimal_mark` to accomodate different numeric styles,
`date_names` to format the date names according to the tradition in a specific location,
and `encoding` to specify the encoding used by the file.
I think `tz` for time zone would also be useful.

1.  What happens if you try and set `decimal_mark` and `grouping_mark` 
    to the same character? What happens to the default value of 
    `grouping_mark` when you set `decimal_mark` to ","? What happens
    to the default value of `decimal_mark` when you set the `grouping_mark`
    to "."?

```{r, error=TRUE}
locale(decimal_mark = ".", grouping_mark = ".")
locale(decimal_mark = ",")$grouping_mark
locale(grouping_mark = ".")$decimal_mark
```

1.  I didn't discuss the `date_format` and `time_format` options to
    `locale()`. What do they do? Construct an example that shows when 
    they might be useful.

The `date_format` can be used to parse dates that are not in the default
`YYYY-MM-DD` format:

```{r}
parse_date("01/31/2000")
# January 31, 2000
parse_date("01/31/2000", locale = locale(date_format = "%m/%d/%Y"))
```

According to the [readr locales
vignette](https://cran.r-project.org/web/packages/readr/vignettes/locales.html),
the argument `time_format` is not used, so it is never useful. But the vignette
is outdated. `time_format` is used exactly the same as `date_format`.

```{r}
parse_time("17:55:14")
parse_time("5:55:14 PM")
# Example of a non-standard time
parse_time("h5m55s14 PM")
parse_time("h5m55s14 PM", locale = locale(time_format = "h%Hm%Ms%S %p"))
```

1.  If you live outside the US, create a new locale object that encapsulates 
    the settings for the types of file you read most commonly.

You can create it by passing custom arguments to `locale` and saving the result.
Many languages are already supported:

```{r}
(es <- locale("es"))
str(es)
```
    
1.  What's the difference between `read_csv()` and `read_csv2()`?

> read_csv2() uses ; for the field separator and , for the decimal point. This
is common in some European countries.

1.  What are the most common encodings used in Europe? What are the
    most common encodings used in Asia? Do some googling to find out.

From the online book [Programming with
Unicode](https://unicodebook.readthedocs.io/) (CC BY-SA 3.0
[license](https://unicodebook.readthedocs.io/about.html#license)), the most
[popular
encodings](https://unicodebook.readthedocs.io/encodings.html#popularity) on the
internet are:

```
1st (56%): ASCII
2nd (23%): Western Europe encodings (ISO 8859-1, ISO 8859-15 and cp1252)
3rd (8%): Chinese encodings (GB2312, …)
and then come Korean (EUC-KR), Cyrillic (cp1251, KOI8-R, …), East Europe (cp1250, ISO-8859-2), Arabic (cp1256, ISO-8859-6), etc.
(UTF-8 was not used on the web in 2001)
```

Note that I used [DuckDuckGo](https://duckduckgo.com/) for the online search :-)

1.  Generate the correct format string to parse each of the following 
    dates and times:

See `?strptime` for the available conversion specifiers (not sure whether to be
proud or depressed that I remembered off the top of my head that `%B` was the
full month name).

```{r}
d1 <- "January 1, 2010"
parse_date(d1, "%B %d, %Y")
# Alternatively can specify date_format via locale argument
parse_date(d1, locale = locale(date_format = "%B %d, %Y"))
d2 <- "2015-Mar-07"
parse_date(d2, "%Y-%b-%d")
d3 <- "06-Jun-2017"
parse_date(d3, "%d-%b-%Y")
d4 <- c("August 19 (2015)", "July 1 (2015)")
parse_date(d4, "%B %d (%Y)")
d5 <- "12/30/14" # Dec 30, 2014
parse_date(d5, "%m/%d/%y")
t1 <- "1705"
parse_time(t1, "%H%M")
t2 <- "11:15:10.12 PM"
parse_time(t2, "%H:%M:%OS %p")
# Alternatively can specify time_format via locale argument
parse_time(t2, locale = locale(time_format = ("%H:%M:%OS %p")))
```

`%OS` is strange. Apparently it is R-specific, and I couldn't get readr to
accept the decimal argument:

> Specific to R is %OSn, which for output gives the seconds truncated to 0 <= n
<= 6 decimal places (and if %OS is not followed by a digit, it uses the setting
of getOption("digits.secs"), or if that is unset, n = 0). Further, for strptime
%OS will input seconds including fractional seconds. Note that %S does not read
fractional parts on output.
