---
title: "Tidy data with tidyr"
author: "John Blischak"
date: "2019-10-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/tidy-data.html)
* [Source](https://github.com/hadley/r4ds/blob/master/tidy.Rmd)

## Setup

```{r setup, message=FALSE}
library(tidyverse)
```

## Tidy data

p. 151

1.  Using prose, describe how the variables and observations are organised in
    each of the sample tables.

* `table1` - Each observation of a country in a given year is in its own row.
The number of cases and the total population are columns.
* `table2` - This is a longer format. Each observation has two rows, one for
the cases and one for the population.
* `table3` - Each observation of a country in a given year is in its own row.
The number of cases and the total population are combined into a single column.
* `table4a` - Each country is a row. The number of cases in 1999 and 2000 are in
separate columns.
* `table4b` - Each country is a row. The total population in 1999 and 2000 are
in separate columns.

1.  Compute the `rate` for `table2`, and `table4a` + `table4b`. 
    You will need to perform four operations:

    1.  Extract the number of TB cases per country per year.
    1.  Extract the matching population per country per year.
    1.  Divide cases by population, and multiply by 10000.
    1.  Store back in the appropriate place.
    
    Which representation is easiest to work with? Which is hardest? Why?

The example tables are included as data sets when the tidyr package is loaded
(see `?table1`).

```{r table2}
tb <- table2 %>% filter(type == "cases") %>% select(country, year, tb = count)
pop <- table2 %>% filter(type == "population")  %>% select(pop = count)
rate <- tb$tb / pop$pop * 10^4
cbind(tb, pop, rate)
```

```{r table4}
tb <- c(table4a$`1999`, table4a$`2000`)
pop <- c(table4b$`1999`, table4b$`2000`)
rate <- tb / pop * 10^4
cbind(table4a$country, tb, pop, rate)
```

They were both kind of awkward.

1.  Recreate the plot showing change in cases over time using `table2`
    instead of `table1`. What do you need to do first?

Need to convert from long to wide. Kind of putting the cart before the horse.
I guess maybe they want you to think about how painful this is to do manually.

I'll reuse my code from above:

```{r cases-over-time}
tb <- table2 %>% filter(type == "cases") %>% select(country, year, tb = count)
pop <- table2 %>% filter(type == "population")  %>% select(pop = count)
rate <- tb$tb / pop$pop * 10^4
cbind(tb, pop, rate) %>%
  ggplot(aes(x = year, y = tb)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country))
```

## Spreading and gathering

p. 156

1.  Why are `gather()` and `spread()` not perfectly symmetrical?  
    Carefully consider the following example:
    
    ```{r, eval=TRUE}
    stocks <- tibble(
      year   = c(2015, 2015, 2016, 2016),
      half  = c(   1,    2,     1,    2),
      return = c(1.88, 0.59, 0.92, 0.17)
    )
    stocks %>% 
      spread(year, return) %>% 
      gather("year", "return", `2015`:`2016`)
    ```
    
    (Hint: look at the variable types and think about column _names_.)

Year is converted to a character vector after `gather()` moves them from
column names to their own column.

    Both `spread()` and `gather()` have a `convert` argument. What does it 
    do?

If `convert=TRUE` (the default is `FALSE`), then `utils::type.convert()` will
be run on the newly created columns. This results in `year` being converted to
an integer, which is still different from its original status as a numeric vector.

```{r}
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`, convert = TRUE)
```


1.  Why does this code fail?

    ```{r, error = TRUE}
    table4a %>% 
      gather(1999, 2000, key = "year", value = "cases")
    ```

Because `1999` and `2000` are numbers, they are interpreted as indicating the
columns to gather by their position. Using column position is an alternative way
to specify the columns:

```{r}
table4a %>%
  gather(2, 3, key = "year", value = "cases")
```

1.  Why does spreading this tibble fail? How could you add a new column to fix
    the problem?

    ```{r}
    people <- tribble(
      ~name,             ~key,    ~value,
      #-----------------|--------|------
      "Phillip Woods",   "age",       45,
      "Phillip Woods",   "height",   186,
      "Phillip Woods",   "age",       50,
      "Jessica Cordero", "age",       37,
      "Jessica Cordero", "height",   156
    )
    ```

It fails because there are two entries for name "Phillip Woods" and "age".
This could be fixed by adding a new column with a unique ID for each individual,
so that people with the same name could be distinguished.

```{r error=TRUE}
people %>%
  spread(key, value)
```

1.  Tidy the simple tibble below. Do you need to spread or gather it?
    What are the variables?

    ```{r}
    preg <- tribble(
      ~pregnant, ~male, ~female,
      "yes",     NA,    10,
      "no",      20,    12
    )
    ```

The variables are pregnancy status (yes, no), gender (male, female), and
count (the number of observations of each combination of pregnancy status and
gender). It needs to be gathered so that the variable gender is its own column.

```{r}
preg %>%
  gather(key = "gender", value = "count", male, female)
```

## Separating and uniting

p. 160

Note in the print copy I have, the section is mistakenly named "Separating and
pull".

1.  What do the `extra` and `fill` arguments do in `separate()`? 
    Experiment with the various options for the following two toy datasets.
    
    ```{r, eval = FALSE}
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"))
    
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"))
    ```

The argument `extra` "controls what happens when there are too many pieces".

```{r extra}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j"))
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"))
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "drop")
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")
```

The argument `fill` "controls what happens when there are not enough pieces".

```{r fill}
tibble(x = c("a,b,c", "d,e", "f,g,i"))
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")
```

1.  Both `unite()` and `separate()` have a `remove` argument. What does it
    do? Why would you set it to `FALSE`?

The argument `remove` removes the original column(s) that was/were separated or
united from the final output. You would set it to `FALSE` if you wanted to
maintain these columns for your analysis. For example, it could be useful to
have the 4-digit year, 2-digit year, and century in the same data frame.

1.  Compare and contrast `separate()` and `extract()`.  Why are there
    three variations of separation (by position, by separator, and with
    groups), but only one unite?

`extract()` uses capture groups from a regular expression. `separate()` uses a
character or numeric position to split.

Because there are so many potential ways to separate a character string. On
the other hand, uniting only requires the connecting character. And if you
wanted to unite from subsets of the existing columns, you could first run
`separate()` to create the new columns and then unite them.


## Missing values

p.163

1.  Compare and contrast the `fill` arguments to `spread()` and `complete()`. 

The argument `fill` to both functions provides the option of specifying the value
to use for missing values (instead of `NA`). The difference is that `spread()`
only accepts one argument; whereas, `complete()` expects a named list where
the name corresponds to the column where the missing value should be replaced.

1.  What does the direction argument to `fill()` do?

Determines whether to fill in missing values from top to bottom (`"down"`, the
default) or from bottom to top (`"up"`). There are also the options `"downup"`
and `"updown"`, but it is unclear to me how these work from the description, and
the toy data set doesn't reveal any difference with their shorter counterparts.

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment %>% 
  fill(person, .direction = "down")
treatment %>% 
  fill(person, .direction = "up")
treatment %>% 
  fill(person, .direction = "downup")
treatment %>% 
  fill(person, .direction = "updown")
```

## Case study

p. 168

```{r}
tidywho <- who %>%
  gather(key, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

1.  In this case study I set `na.rm = TRUE` just to make it easier to
    check that we had the correct values. Is this reasonable? Think about
    how missing values are represented in this dataset. Are there implicit
    missing values? What's the difference between an `NA` and zero? 

The data set contains both `NA` and zeros, e.g.

```{r}
length(who$new_sp_f014)
sum(is.na(who$new_sp_m014))
sum(who$new_sp_m014  == 0, na.rm = TRUE)
```

```{r}
tidywho_na <- who %>%
  gather(key, value, new_sp_m014:newrel_f65, na.rm = FALSE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

1.  What happens if you neglect the `mutate()` step?
    (`mutate(key = stringr::str_replace(key, "newrel", "new_rel"))`)

Because there aren't enough pieces (2 instead of 3), all the variables with
`newrel` have their `sexage` stored in `var`, which results in `sex` and `age`
being missing for all the new cases of relapsed TB.

```{r}
 who %>%
  gather(key, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  separate(key, c("new", "var", "sexage")) %>%
  tail
```

1.  I claimed that `iso2` and `iso3` were redundant with `country`. 
    Confirm this claim.

```{r}
who %>%
  group_by(country) %>%
  summarize(n_iso2 = n_distinct(iso2),
            n_iso3 = n_distinct(iso3)) %>%
  {stopifnot(.$n_iso2 == 1, .$n_iso3 == 1)}
# Note: need to surround last expression in {} because this prevents . from
# being passed as the first argument to stopifnot(), which obviously causes a
# failure since a tibble does not evaluate to TRUE
# https://stackoverflow.com/a/42386886/2483477
```

1.  For each country, year, and sex compute the total number of cases of 
    TB. Make an informative visualisation of the data.

```{r}
tidywho %>%
  group_by(country, year, sex) %>%
  summarize(total_cases = sum(value)) %>%
  ggplot(aes(x = year, y = total_cases)) +
  geom_point(aes(color = country)) +
  geom_smooth() +
  facet_wrap(~sex) +
  theme(legend.position = "none")
```
