---
title: "Relational data with dplyr"
author: "John Blischak"
date: "2019-10-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

* [Book chapter](https://r4ds.had.co.nz/relational-data.html)
* [Source](https://github.com/hadley/r4ds/blob/master/relational-data.Rmd)

## Setup

```{r setup, message=FALSE}
library(tidyverse)
library(nycflights13)
library(Lahman)
library(babynames)
library(nasaweather)
library(fueleconomy)
```

## nycflights13

p. 174

1.  Imagine you wanted to draw (approximately) the route each plane flies from
    its origin to its destination. What variables would you need? What tables
    would you need to combine?

I would need the variables `origin` and `dest` from `flights` and would need to
combine these with the `lat` and `lon` from `airports`.

1.  I forgot to draw the relationship between `weather` and `airports`.
    What is the relationship and how should it appear in the diagram?

The variable `origin` in `weather` corresponds to the variable `faa` in
`airports`.

1.  `weather` only contains information for the origin (NYC) airports. If
    it contained weather records for all airports in the USA, what additional
    relation would it define with `flights`?

It would also define a relation with `dest`.

1.  We know that some days of the year are "special", and fewer people than
    usual fly on them. How might you represent that data as a data frame?
    What would be the primary keys of that table? How would it connect to the
    existing tables?

I could create a table `special` that contained the variables `name`, `month`,
and `day` for each "special" day of the year. The variables `month` and `day`
would connect to the variables of the same name in `flights` and `weather`.

## Keys

p. 177

1.  Add a surrogate key to `flights`.

```{r}
flights %>%
  mutate(surrogate = row_number()) %>%
  select(surrogate, everything())
```

1.  Identify the keys in the following datasets

    1.  `Lahman::Batting`,
    1.  `babynames::babynames`
    1.  `nasaweather::atmos`
    1.  `fueleconomy::vehicles`
    1.  `ggplot2::diamonds`
    
    (You might need to install some packages and read some documentation.)

```{r}
head(Lahman::Batting)
?Lahman::Batting
# playerID, yearID, stint
Lahman::Batting %>%
  count(playerID, yearID, stint) %>%
  filter(n > 1)
babynames::babynames
?babynames::babynames
# year, sex, name
babynames::babynames %>%
  count(year, sex, name) %>%
  filter(n > 1)
nasaweather::atmos
?nasaweather::atmos
# lat, long, year, month
nasaweather::atmos %>%
  count(lat, long, year, month) %>%
  filter(n > 1)
fueleconomy::vehicles
?fueleconomy::vehicles
# id
fueleconomy::vehicles %>%
  count(id) %>%
  filter(n > 1)
ggplot2::diamonds
?ggplot2::diamonds
# none
ggplot2::diamonds %>%
  select_if(is.factor) %>%
  count(cut, color, clarity) %>%
  filter(n > 1)
```

1.  Draw a diagram illustrating the connections between the `Batting`,
    `Master`, and `Salaries` tables in the Lahman package. Draw another diagram
    that shows the relationship between `Master`, `Managers`, `AwardsManagers`.

    How would you characterise the relationship between the `Batting`,
    `Pitching`, and `Fielding` tables?

## Mutating joins

p. 186


1.  Compute the average delay by destination, then join on the `airports`
    data frame so you can show the spatial distribution of delays. Here's an
    easy way to draw a map of the United States:

    ```{r, eval = TRUE}
    airports %>%
      semi_join(flights, c("faa" = "dest")) %>%
      ggplot(aes(lon, lat)) +
        borders("state") +
        geom_point() +
        coord_quickmap()
    ```

    (Don't worry if you don't understand what `semi_join()` does --- you'll
    learn about it next.)

    You might want to use the `size` or `colour` of the points to display
    the average delay for each airport.

```{r delay-by-dest-map}
flights %>%
  group_by(dest) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airports, c("dest" = "faa")) %>%
  ggplot(aes(lon, lat)) +
  borders("state") +
  geom_point(aes(color = avg_delay)) +
  coord_quickmap()
```

1.  Add the location of the origin _and_ destination (i.e. the `lat` and `lon`)
    to `flights`.

```{r}
flights %>%
  left_join(airports, by = c("origin" = "faa")) %>%
  left_join(airports, by = c("dest" = "faa"), suffix = c(".origin", ".dest")) %>%
  select(contains("origin"), contains("dest"))
```

1.  Is there a relationship between the age of a plane and its delays?

```{r}
age_delay <- flights %>%
  left_join(planes, by = "tailnum", suffix = c(".flight", ".plane"))
ggplot(age_delay, aes(x = year.plane, y = arr_delay)) +
  geom_point() +
  geom_smooth()
lm(arr_delay ~ year.plane, data = age_delay)
```

1.  What weather conditions make it more likely to see a delay?

```{r}
flights_weather <- flights %>%
  left_join(weather)
lm(arr_delay ~ dewp + humid + wind_dir + wind_speed + wind_gust + precip + pressure + visib,
   data = flights_weather)
```

1.  What happened on June 13 2013? Display the spatial pattern of delays,
    and then use Google to cross-reference with the weather.

```{r}
weather %>%
  filter(year == 2013, month == 6, day == 13)
```

```{r 2013-06-13}
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarize(avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airports, c("dest" = "faa")) %>%
  ggplot(aes(lon, lat)) +
  borders("state") +
  geom_point(aes(color = avg_delay)) +
  coord_quickmap()
```

https://en.wikipedia.org/wiki/June_12%E2%80%9313,_2013_derecho_series

## Filtering joins

p. 191

1.  What does it mean for a flight to have a missing `tailnum`? What do the
    tail numbers that don't have a matching record in `planes` have in common?
    (Hint: one variable explains ~90% of the problems.)

Not sure

```{r}
flights %>%
  anti_join(planes, by = "tailnum")

flights %>%
  anti_join(planes, by = "tailnum") %>%
  filter(nchar(tailnum) == 6)

flights %>%
  semi_join(planes, by = "tailnum") %>%
  filter(nchar(tailnum) == 5)
```

1.  Filter flights to only show flights with planes that have flown at least 100
    flights.

```{r}
planes_at_least_100 <- flights %>%
  group_by(tailnum) %>%
  summarize(n_flights = n()) %>%
  filter(n_flights >= 100)
flights_at_least_100 <- flights %>%
  semi_join(planes_at_least_100, by = "tailnum")
dim(flights_at_least_100)
```

1.  Combine `fueleconomy::vehicles` and `fueleconomy::common` to find only the
    records for the most common models.

```{r}
fueleconomy::vehicles %>%
  semi_join(fueleconomy::common)
```

1.  Find the 48 hours (over the course of the whole year) that have the worst
    delays. Cross-reference it with the `weather` data. Can you see any
    patterns?

1.  What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you?
    What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?

`anti_join(flights, airports, by = c("dest" = "faa"))` tells you the flights
with a destination that isn't recorded in `airports`.

```{r}
anti_join(flights, airports, by = c("dest" = "faa"))
```

`anti_join(airports, flights, by = c("faa" = "dest"))` tells you the airports
that were not flown to directly from NYC airports.

1.  You might expect that there's an implicit relationship between plane
    and airline, because each plane is flown by a single airline. Confirm
    or reject this hypothesis using the tools you've learned above.

```{r}
flights %>%
  group_by(tailnum) %>%
  summarize(n_airlines = n_distinct(carrier)) %>%
  filter(n_airlines != 1)
```
